Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO

### 2015-11-09 Paper/paper.tex
multi-row contour plots now workcd ../..

\usepackage[colorlinks, pdfborder={0 0 0}, plainpages=false]{hyperref}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage[usenames,dvipsnames]{color}
\usepackage{amssymb}

@@

% Macros for text changes
\newcommand{\red}{\textcolor{red}}
\newcommand{\dan}[1]{\textcolor{WildStrawberry}{#1}}
\newcommand{\Mark}[1]{\textcolor{Cerulean}{#1}}
\newcommand{\larry}[1]{\textcolor{OliveGreen}{#1}}
\newcommand{\bela}[1]{\textcolor{Blue}{#1}}
\newcommand{\saul}[1]{\textcolor{Orange}{#1}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Current status:

\begin{enumerate}
\item We choose the following values of different NS-BH binary parameters:
\begin{enumerate}
\item $m_\mathrm{NS}=1.35M_\odot$;
\item $q=m_\mathrm{BH}/\mathrm{NS}=\{2,3\}$;
\item $\chi_\mathrm{NS}=0$;
\item $\chi_\mathrm{BH}=\{-0.5, 0, +0.5\}$;
\item $\Lambda =\{500,1000,2000\}$.
\end{enumerate}

\item For each system comprised of all unique combinations of the above 
parameter choices, we carry out three parameter estimation tests: 
\begin{enumerate}
\item BH-BH signals injected, recovered with BH-BH waveforms (control),
\item NS-BH signals injected, recovered with BH-BH waveforms, (current approach) and
\item NS-BH signals injected, recovered with NS-BH waveforms.
\end{enumerate}
Injections were made at different SNR values: 
\begin{enumerate}
\item $\rho = \{20, 30, 50, 70, 90, 120\}$.
\end{enumerate}
Choice of emcee parameters:
\begin{enumerate}
\item $N_\mathrm{samples}=150,000$; 
\item $N_\mathrm{walkers}=100$; 
\item $N_\mathrm{burn-in}=500$; 
\end{enumerate}
and choice of prior boundaries:
\begin{enumerate} 
\item $m_\mathrm{BH}\in [1.2, 25]M_\odot$;
\item $m_\mathrm{NS}\in [1.2, 15]M_\odot$;
\item $\Lambda\leq 4000$; 
\item $\sigma\left(\Lambda\right) = 100$ (for chains where templates are NS-BH waveforms).
\end{enumerate}

@@

\begin{enumerate}
\item Above what SNR values, below what mass-ratios, above what BH spins, etc, 
do we begin to care about NS deformability?\newline
``one interesting plot to make would be the parameter bias (max-Likelihood
parameters minus true parameters) vs SNR, as functionals of q/sBH/Lambda, when
using T signals recovered with N templates.''
\end{enumerate}

>>>>

\usepackage[colorlinks, pdfborder={0 0 0}, plainpages=false]{hyperref}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage[usenames,dvipsnames]{color}
\usepackage{amssymb}
\usepackage[dvipsnames]{xcolor}

@@

% Macros for text changes
\newcommand{\red}{\textcolor{red}}
\newcommand{\dan}[1]{\textcolor{WildStrawberry}{#1}}
\newcommand{\Mark}[1]{\textcolor{Cerulean}{#1}}
\newcommand{\larry}[1]{\textcolor{OliveGreen}{#1}}
\newcommand{\bela}[1]{\textcolor{Blue}{#1}}
\newcommand{\saul}[1]{\textcolor{Orange}{#1}}
\newcommand{\prayush}{\textcolor{red!40!black}}

@@

\newcommand{\NS}{\mathrm{NS}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The Advanced LIGO (aLIGO) observatories began their first observing run ``O1'' in 
September, 2015. With both detectors a factor of $3-4$ more sensitive than
their first-generation counterparts, we have access to $\approx 50$ times more 
spacial volume of the universe at every instant. Within a 
$3$ month observing run, these detectors would have access to as
many binary mergers as initial-LIGO detectors did in $\sim 6$ years of 
coincident operation. Through gradual updates, both of these instruments will
reach their design sensitivity by $2019$, at which point they will like detect
around $70$ binary mergers a year~\cite{Abadie:2010cf}.

Coalescing binaries containing a neutron star (NS) and a black hole (BH) are
one of the most important sources of gravitational waves (GW) with the aLIGO
and Virgo instruments, which expect to detect $\mathcal{O}(10)$
of such systems every 
year~\cite{Abadie:2010cf}. Conventional astronomical methods have, to date, made 
observations of NSs with masses between $XX-YY M_\odot$ and spin
$|\vec{\chi}\lesssim 0.05$. On the other hand, indirect observations of BHs 
have placed the masses of stellar-collapse BHs between $5-35M_\odot$, with 
spin angular momenta ranging from low to nearly extremal values (i.e., nearly
as fast as possible - see,
e.g.~\cite{McClintockEtAl:2006,Miller:2009cw,Gou:2014una} for examples of 
nearly extremal estimates of BH spins, and see
Refs.~\cite{McClintock:2013vwa,Reynolds:2013qqa} for recent reviews of
astrophysical BH spin measurements). \prayush{[Do we want to emphasize on BH
spins here?]}.

One question that aLIGO is likely to probe is `what is the nature of
matter at nuclear densities', i.e. what is the equation of state (EoS) of 
neutron stars. The EoS determines the compactness of NSs, which
determines how early in the coalescence will the NS be tidally disrupted
by the gravitational field of the BH, resulting in the formation
of an accretion disk and/or ejection of unbound matter. This makes NSBH
mergers are one of the prime targets of the program for joint detection of
GW and electromagnetic signals. 
% 
The difficulty in discerning the matter effects in an NSBH merger GW signal
lies in distinguishing it reliably from a BHBH merger signal. The earlier the
NS disrupts, the more orbits there are over which the effect of NS disruption
is visible. Where exactly does disruption happen is governed both by the 
mass-ratio of the binary, as well as the spin on the BH. Larger BHs (with
$q=m_\mathrm{BH}/m_\mathrm{NS}\gtrsim 7$), as well as BHs with large retrograde
spins ($\chi_\mathrm{BH} \lesssim 0$), tend to swallow the NS whole due to a
rapid merger (i.e. without it disrupting before the inner-most stable circular
orbit (ISCO) is reached). On the other hand, binaries with mass-ratios 
$q=2-5$ with large pro-grade $\chi_\mathrm{BH}$ can disrupt much before 
merger, leaving their imprint in the GW signal emitted close to $1.7-2$~kHz.
%

The goal of this paper is to probe the distinguishability of BHBH and NSBH
mergers, focusing at the effects of the BH mass, BH spin, and NS tidal 
deformability. We also investigate the accuracy with which aLIGO can measure
the tidal deformability parameter $\Lambda_\NS\sim(R/M)^5_\NS$. For both of 
these studies, we will use the zero-detuning high-power noise curve for the
aLIGO detectors. \prayush{In addition, we also study the effect of tuning aLIGO
over a narrow frequency range aimed at NSBH mergers.}
We use the models published in Ref.~\cite{Lackey:2013axa,Pannarale:2015jka} to
simulate NSBH merger waveforms.

\vspace{2cm}

\textit{\prayush{Current status:}}

Below we explain the outline the different calculations performed in this 
paper:

\begin{enumerate}
\item We choose the following values of different NSBH binary parameters:
\begin{enumerate}
\item $m_\mathrm{NS}=1.35M_\odot$;
\item $q=m_\mathrm{BH}/\mathrm{NS}=\{2,3\}$;
\item $\chi_\mathrm{NS}=0$;
\item $\chi_\mathrm{BH}=\{-0.5, 0, +0.5\}$;
\item $\Lambda =\{500,1000,2000\}$ ($\Lambda=0\implies \mathrm{BHBH}$).
\end{enumerate}

\item For each system, corresponding to one of all possible unique combinations
of the above parameter choices, we carry out three parameter estimation tests: 
\begin{enumerate}
\item BH-BH signals injected, recovered with BH-BH waveforms (control),
\item NS-BH signals injected, recovered with BH-BH waveforms, (current approach) and
\item NS-BH signals injected, recovered with NS-BH waveforms.
\end{enumerate}
Injections were made at different SNR values: 
\begin{enumerate}
\item $\rho = \{20, 30, 50, 70, 90, 120\}$.
\end{enumerate}
Choice of emcee parameters:
\begin{enumerate}
\item $N_\mathrm{samples}=150,000$; 
\item $N_\mathrm{walkers}=100$; 
\item $N_\mathrm{burn-in}=500$; 
\end{enumerate}
and choice of prior boundaries:
\begin{enumerate} 
\item $m_\mathrm{BH}\in [1.2, 25]M_\odot$;
\item $m_\mathrm{NS}\in [1.2, 15]M_\odot$;
\item $\Lambda\leq 4000$; 
\item $\sigma\left(\Lambda\right) = 100$ (for chains where templates are NS-BH waveforms).
\end{enumerate}

@@

\begin{enumerate}
\item Above what SNR values, below what mass-ratios, above what BH spins, etc, 
do we begin to care about NS deformability?\newline
``one interesting plot to make would be the parameter bias (max-Likelihood
parameters minus true parameters) vs SNR, as functionals of q/sBH/Lambda, when
using T signals recovered with N templates.''\newline
`` another interesting plot to make would be the fitting-factors recovered with
a discrete (search oriented) template bank of BBH waveforms, for NSBH signals of
loudness, deformability, spins, mass-ratios, etc. Which parameter plays the 
dominant role in paramter regions where the bank does not suffice to recover
the desired threshold of optimal SNR?''
\end{enumerate}
